Aligning species range data to better serve science and conservation

### 2016-01-14 data_explore/data_prep.Rmd

```{r get_spp_list, echo = FALSE, warning = FALSE, message = FALSE, eval = TRUE}
### get spp_all_complete from git-annex, bring to local repo.  The spp_all_complete
### file was created in ohiprep/globalprep/SPP_ICO/data_prep_SPP.R, running the
### function create_spp_master_lookup(source_pref = 'iucn', fn_tag = '_complete', reload = TRUE)
spp_all_remote <- read.csv(file.path(dir_anx, scenario, 'int/spp_all_complete.csv'), 
                           stringsAsFactors = FALSE)
write.csv(spp_all_remote, file.path(dir_data, 'spp_all.csv'), row.names = FALSE)

@@

data_file <- file.path(dir_data, 'spp_list_w_area_all.csv')
reload = TRUE
if (!file.exists(data_file) | reload) {
  ### species info list - scinames, IUCN species ID, AM species ID, category etc
  spp_list_base <- read.csv(file.path(dir_data, 'spp_am_v_iucn.csv'),
                            stringsAsFactors = FALSE)

>>>>

```{r get_spp_list, echo = FALSE, warning = FALSE, message = FALSE, eval = FALSE}
### get spp_all_complete from git-annex, bring to local repo.  The spp_all_complete
### file was created in ohiprep/globalprep/SPP_ICO/data_prep_SPP.R, running the
### function create_spp_master_lookup(source_pref = 'iucn', fn_tag = '_complete', reload = TRUE)
spp_all_remote <- read.csv(file.path(dir_anx, scenario, 'int/spp_all.csv'), 
                           stringsAsFactors = FALSE) %>%
  filter(!str_detect(spatial_source, 'subpop')) %>%
  filter(!str_detect(spatial_source,  'alias')) %>%
  select(am_sid, sciname, reviewed, iucn_sid, popn_trend,
         popn_category, spp_group, id_no, spatial_source)

write.csv(spp_all_remote, file.path(dir_data, 'spp_all.csv'), row.names = FALSE)

@@

data_file <- file.path(dir_data, 'spp_list_w_area_all.csv')
reload <- TRUE
if (!file.exists(data_file) | reload) {
  ### species info list - scinames, IUCN species ID, AM species ID, category etc
  spp_list_base <- read.csv(file.path(dir_data, 'spp_am_v_iucn.csv'),
                            stringsAsFactors = FALSE)

### 2016-01-14 ms_1_abst_sig.Rmd

Importantly, biases in taxonomic or spatial coverage within a dataset could shift management or conservation actions away from places or species that are most in need.(maybe Jetz 2008?)  Inaccurate indications of presence or absence could lead to inefficient or ineffective marine reserve systems and management plans. (Rondinini et al. 2006) (Jetz 2008?)
-->
Together, the two datasets provide range information for XXX species, with only `r nrow(spp_map_pairs)` species common to both datasets.  ___what is the key takeaway for the taxonomic and spatial coverage analyses? can we say something about benefits of combining the two for better taxo coverage?___  For the subset of species included in both datasets, we examine the alignment of range maps by comparing range distribution and extent.  Categorizing the results based on quality of alignment, we identify several _problems?_ that commonly lead to disagreement in species range predictions.  Notably, IUCN maps often predict species presence at unsuitable depths, and AquaMaps maps often extrapolate species presence far afield from known occurrences.  ___something about recommendations here? how this understanding can lead to better use of these datasets?___

To understand the implications of these differences, we reexamined two recent global analyses that depend on these datasets: the Ocean Health Index and a global analysis of gaps in coverage of marine protected areas.  ___how far to describe results here?___
<!--
To understand the implications of differences between these two datasets, we compared how each data set represents the global spatial and taxonomic distribution of species.  For the relatively small number of species mapped in both datasets, we examined how well the species maps align. We then subjected two recent marine biodiversity studies - the Ocean Health Index biodiversity goal (Halpern et al. 2012) and a global MPA gap analysis (Klein et al. 2015) - to a sensitivity analysis, substituting one dataset over the other, to highlight the possible consequences of different data use decisions on our understanding of _the status of_ marine biodiversity.
-->

>>>>

Importantly, biases in taxonomic or spatial coverage within a dataset could shift management or conservation actions away from places or species that are most in need.(maybe Jetz 2008?)  Inaccurate indications of presence or absence could lead to inefficient or ineffective marine reserve systems and management plans. (Rondinini et al. 2006) (Jetz 2008?)
-->
Together, the two datasets provide range information for XXX species, with only `r nrow(spp_map_pairs)` species common to both datasets.  ___what is the key takeaway for the taxonomic and spatial coverage analyses? can we say something about benefits of combining the two for better taxo coverage?___

For the subset of species included in both datasets, we examine the alignment of range maps by comparing distribution and extent of predicted ranges.  Categorizing the results based on quality of alignment, we identify several issues that commonly lead to disagreement in species range predictions.  Notably, IUCN maps often neglect considerations of bathymetry, leading to predictions of species presence at unsuitable depths, and AquaMaps maps for data-poor species often extrapolate presence far afield from known occurrences.  ___something about recommendations here? how this understanding can lead to better use of these datasets?___

To understand the implications of these differences, we reexamined two recent global analyses that depend on these datasets: the Ocean Health Index and a global analysis of gaps in coverage of marine protected areas.  Variations in which dataset is used, and how, lead to significantly different estimates of species diversity and conservation effectiveness.

<!--
To understand the implications of differences between these two datasets, we compared how each data set represents the global spatial and taxonomic distribution of species.  For the relatively small number of species mapped in both datasets, we examined how well the species maps align. We then subjected two recent marine biodiversity studies - the Ocean Health Index biodiversity goal (Halpern et al. 2012) and a global MPA gap analysis (Klein et al. 2015) - to a sensitivity analysis, substituting one dataset over the other, to highlight the possible consequences of different data use decisions on our understanding of _the status of_ marine biodiversity.
-->

### 2016-01-14 ms_2_intro_res_disc.Rmd

For `r n_q2_am_sm` of `r n_q2` species in quadrant 2 (`r round((n_q2_am_sm/n_q2) * 100, 1)`%), the IUCN extent is larger than the AquaMaps extent.  By itself, this is not surprising; but for many of these species, a quick look at the maps (see SOM for examples) shows that the IUCN range hews closely to the AquaMaps range, while adding a significant buffer zone.  Coral species dominate this quadrant (`r n_q2_corals` coral species - `r round(n_q2_corals/n_corals * 100, 1)`% of all corals and `r round(n_q2_corals/n_q2 * 100, 1)`% of all species in quadrant 2).  We suspect that many of these extent-misaligned map pairs can be easily explained: most corals and reef-associated organisms prefer shallower waters; seafloor depth is explicitly modeled in AquaMaps, but not explicitly considered in IUCN range considerations. (For examples and details, see SOM)

@@

The extent-alignment of species maps found in quadrant 3 can in many cases be attributed to "two wrongs make a right."  For these species, IUCN ranges are frequently overextended into unsuitable habitat, as in the case of many quadrant 2 species, but at the same time, AquaMaps aggressively extrapolates distribution of presence into locations where IUCN predicts absence.  This results in an area ratio close to 100%, giving the impression of extent alignment. ___this quadrant seems to be a false alignment then?___

Quadrant 4 species maps seem to suffer from the same problem as those in quadrant 3, but here the AquaMaps extrapolation extends even further, pushing the extent alignment ratio lower.  In this quadrant, AquaMaps predicts a larger extent than IUCN for 70% of species represented in this quadrant, compared with just 30% of species in the entire set of overlapping maps _(this is supposed to show that much of this quadrant is based on overextrapolation of AquaMaps)_.

___table to show data poor status and reviewed status for AquaMaps maps represented in the quadrant plot___

| quadrant | n species | n data poor species | n reviewed species |
| :------: | :-------: | :-----------------: | :----------------: |
|  total   |   22889   |        8749 (38.2%) |       1296  (5.7%) |
|  paired  |    2166   |         457 (21.1%) |        290 (13.4%) |
|   q1     |     401   |          33  (8.2%) |        100 (24.9%) |
|   q2     |     682   |         151 (22.1%) |        100 (14.7%) |
|   q3     |     682   |         114 (16.5%) |         65  (9.5%) |
|   q4     |     410   |         159 (39.7%) |         25  (6.2%) |

>>>>

For `r n_q2_am_sm` of `r n_q2` species in quadrant 2 (`r round((n_q2_am_sm/n_q2) * 100, 1)`%), the IUCN extent is larger than the AquaMaps extent.  By itself, this is not surprising; but for many of these species, a quick look at the maps (see SOM for examples) shows that the IUCN range hews closely to the AquaMaps range, while adding a significant buffer zone.  Coral species dominate this quadrant (`r n_q2_corals` coral species - `r round(n_q2_corals/n_corals * 100, 1)`% of all corals and `r round(n_q2_corals/n_q2 * 100, 1)`% of all species in quadrant 2).  We suspect that many of these extent-misaligned map pairs can be easily explained: most corals and reef-associated organisms prefer shallower waters; seafloor depth is explicitly modeled in AquaMaps, while depth is frequently ignored as a factor in IUCN range maps. (For examples and details, see SOM)

@@

The extent-alignment of species maps found in quadrant 3 can in many cases be attributed to "two wrongs make a right."  For these species, IUCN ranges are frequently overextended into unsuitable habitat, as in the case of many quadrant 2 species, while at the same time, AquaMaps aggressively extrapolates distribution of presence into locations where IUCN predicts absence.  This results in an area ratio close to 100%, giving the impression of extent alignment. ___this quadrant seems to be a false alignment then?___

Quadrant 4 species maps seem to suffer from the same problem as those in quadrant 3, but here the AquaMaps extrapolation extends even further, pushing the extent alignment ratio lower.  In this quadrant, AquaMaps predicts a larger extent than IUCN for 70% of species represented in this quadrant, compared with just 30% of species in the entire set of overlapping maps _(this is supposed to show that much of this quadrant is based on overextrapolation of AquaMaps)_. Quadrant 4 in particular includes a high proportion of data-poor species in which the environmental envelope model is based on fewer than 10 known cell occurrences, and at the same time includes a smaller number of reviewed species maps than the other quadrants.

___table to show data poor status and reviewed status for AquaMaps maps represented in the quadrant plot - should the quadrant plot show data-poor species instead of, say, red-list category?___

| quadrant | n species | n data poor species | n reviewed species |
| :------: | :-------: | :-----------------: | :----------------: |
| overall  |   22889   |        8749 (38.2%) |       1296  (5.7%) |
| AM&IUCN  |    2166   |         457 (21.1%) |        290 (13.4%) |
|   q1     |     401   |          33  (8.2%) |        100 (24.9%) |
|   q2     |     682   |         151 (22.1%) |        100 (14.7%) |
|   q3     |     682   |         114 (16.5%) |         65  (9.5%) |
|   q4     |     410   |         159 (39.7%) |         25  (6.2%) |
